---
title: "2025LMMCourse Session 5 Practical"
author: "Fang Yang"
date: "2025-05-16"
output: bookdown::pdf_document2
toc: false
editor_options: 
  chunk_output_type: console
---


```{r install packages, include=FALSE}
library("tidyverse")
library("lme4")
library("effects")
library("sjPlot") # for creating regression models
library("webshot") # for converting tab_model shot from html to png and include it in the .pdf file of the report
library("kableExtra") # for creating descriptive statistics tables
library("patchwork") # for arrange plots 

library("broom.mixed") #"broom" for mixed models, used to turn messy R output into nicer table and plots
library("performance")

knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```


Try to explore syntactic priming effect in the cheese data.

Each time a participant and a confederate took turns to describe pictures to each other. The confederate pretended to be a real participant but read out scripts. A response were coded as 1 if the real participant reused the confederate's sentence structure, otherwise 0. Participants did the experiments online with or without seeing each other's faces (video_call vs audio_call).

## Data loading, cleaning and visualisation

```{r}
data_cheese <- read_csv("dog_man_cheese.csv")
```

```{r}
summary(data_cheese)


data_cheese$Response<-as.factor(data_cheese$Response)   # 1= aligned; 0 = not aligned


data_cheese$prime <- factor(data_cheese$prime, levels=c("A","B")) # set structure A as reference level 

data_cheese$communication <- factor(data_cheese$communication, levels=c("video_call","audio_call")) 
```


```{r, out.width = "80%", out.height = "100%"}

data_cheese |> 
    drop_na() |> 
    ggplot(aes( x=prime, fill=Response)) +
  geom_bar(position = "fill") +
  facet_wrap(~communication)+
  ggtitle(label = "Participants' choice of structure")+
  labs(y = "Proportion", 
       x= "Prime structure" )+ 
  theme(plot.title = element_text(hjust = 0.5, size=18,face = "bold"),
        axis.text.x = element_text(size=14,face = "bold"),
        axis.title.x = element_text(size=16,face = "bold"),
        axis.title.y = element_text(size=16,face = "bold"),
        legend.title = element_text(size=16,face = "bold"),
        legend.text = element_text(size=12,face = "bold")
  )
```

## Fit an additive model

Fit an intercept-only model including the following as fixed effects: (1) main 
effect of prime and (2) main effect of communication. 
 
```{r}
m_addi_datacheese <- glmer(Response ~ prime + communication+ 
                      (1|participant) + 
                      (1|item), 
                    data = data_cheese, 
                    family='binomial',
                    na.action=na.exclude,
                    control=glmerControl(optimizer="bobyqa",optCtrl=list(maxfun=2e5))) 

summary(m_addi_datacheese) 
```

## Fit an interactive model

Fit an intercept-only model including the following as fixed effects: (1) main 
effect of prime, (2) main effect of communication, and (3) the interaction effect 
between prime and communication. 

```{r}
m_int_datacheese <- glmer(Response ~ prime*communication+ 
                      (1|participant) + 
                      (1|item), 
                    data = data_cheese, 
                    family='binomial',
                    na.action=na.exclude,
                    control=glmerControl(optimizer="bobyqa",optCtrl=list(maxfun=2e5))) 

summary(m_int_datacheese) 
```

## Model comparison

```{r}
anova(m_addi_datacheese, m_int_datacheese)
```

## Interpret the results

```{r}
summary(m_int_datacheese)
```

Hint: remember the coefficients are log odds. you need to transfer them to odds and then probability

### Convert log odds to odds

### (1) Probability of producing B in video+primeA condition (intercept)
```{r}
Prob_A_video <- exp(0.43)/(exp(0.43)+1)
Prob_A_video
```

### (2) Probability of producing B in video+primeB condition(intercept + slope)

```{r}
Prob_B_video <- exp(1.26)/(exp(1.26)+1) 
Prob_B_video
```

### (3) Probability of producing B in audio+primeA condition (intercept + slope)
```{r}
Prob_A_audio <- exp(0.53)/(exp(0.53)+1) 
Prob_A_audio
```

### (4)Probability of producing B in audio+primeB condition (intercept + slope)
```{r}
Prob_B_audio <- exp(0.65)/(exp(0.65)+1) 
Prob_B_audio
```

## Visulise the Model

```{r, out.width = "80%", out.height = "100%"}
# fixed effects
plot_model(m_int_datacheese, type = "int")
```

```{r, out.width = "50%", out.height = "80%"}
# random effects 
randoms_cheese <- ranef(m_int_datacheese, condVar=TRUE)
dotplot.ranef.mer(randoms_cheese)
```

## Check and plot random effects

```{r results='hide'}
### to check the random effect
ranef(m_int_datacheese)

### to check the fixed effect
fixef(m_int_datacheese)

### group level coefficient
coef(m_int_datacheese)
```

## Check Model Assumptions

For generalized linear mixed models, returns the QQ-plot for random effects.

```{r, out.width = "50%", out.height = "80%"}
sjPlot::plot_model(m_int_datacheese, type = "diag") 


```


